Search-Based Test Data Generation for JavaScript Functions that Interact with the DOM

### 2018-05-11 doc/icse2018/issta2018.tex
Near final version of the issre 2018 paper

\newcommand{\Server}{\emph{Server}\xspace}
\newcommand{\Client}{\emph{Client}\xspace}
\newcommand{\User}{\emph{User}\xspace}
\newcommand{\Validator}{\emph{Validator}\xspace}
\newcommand{\thenBr}{\text{then}}
\newcommand{\elseBr}{\text{else}}
\newcommand{\inFor}{\text{inFor}}
\newcommand{\inWhile}{\text{while}}
\newcommand{\un}[1]{\underline{#1}}
\newcommand{\Random}{\mathbb{R}}
\newcommand{\Genetic}{\mathbb{G}}
\newcommand{\RGenetic}{\mathbb{G}_{10}}

@@

\begin{abstract}
The popularity of JavaScript is continuously growing. Together with HTML and CSS, it forms the core technology stack for web development. Because of the dynamic nature and complex interplay with HTML, JS applications are still error-prone and vulnerable. Despite the active research efforts to devise intricate static and dynamic analyses to facilitate JS testing, the problem of test data generation for JS applications interacting with the DOM has not yet been addressed. In this paper, we present a novel \textbf{J}avascript \textbf{E}volution-based testing framework with \textbf{D}OM as an \textbf{I}nput, called \emph{JEDI}. In order to reach a target branch, it applies genetic search for relevant input parameters of the JS function including the global DOM state. We conducted an empirical evaluation to study the effectiveness and efficiency of our testing framework. It shows that the \emph{genetic with restart} algorithm, proposed in this paper, is able to achieve complete branch coverage for all experimental subjects, taking on average 19 seconds per branch.
\end{abstract}

@@

\begin{figure}[t]
  \begin{lstlisting}[style=htmlcssjs,language=JavaScript]
/*t dom */
function isGameFinished() {
  var obj = document.getElementById('sudoku'); |c \label{isGameFinished.getSudoku} |c
  var subDivs = obj.getElementsByTagName('DIV'); |c \label{isGameFinished.getDivs} |c
  var allOk = true;
  for (var no = 0; no < subDivs.length; no++) { |c \label{isGameFinished.inFor.begin} |c
    if (subDivs[no].className.indexOf('sudokuSquare') >= 0 |c \label{isGameFinished.if1.begin} |c 
        && !subDivs[no].style.backgroundColor) { 
      var spans = subDivs[no].getElementsByTagName('SPAN');
      if (spans[0].innerHTML != spans[1].innerHTML) { |c \label{isGameFinished.if2.begin} |c
        allOk = false; |c \label{isGameFinished.unfinished} |c
        break; 
      }
    }
  } |c \label{isGameFinished.outFor} |c
}
\end{lstlisting}
  \caption{JS function \texttt{isGameFinished.js} from sudoku}
  \label{code.isGameFinished}
\end{figure}

@@

<!-- T3: (7,8), (10,14) -->   <!-- T4: (7,8), (10,11) -->
<html>                        <html>
 <body>                        <body>
  <div id='sudoku'>             <div id='sudoku'>
   <div class='sudokuSquare'>    <div class='sudokuSquare'>
    <span></span>                 <span></span> 
    <span></span>                 <span>TEST</span>
   </div>                        </div> 
  </div>                        </div>
 </body>                       </body>
</html>                       </html>
  \end{lstlisting}
  \caption{Input DOM arguments for \texttt{isGameFinished.js}}
  \label{fig.isGameFinished.tests}
\end{figure}

Let us consider the JS function \texttt{isGameFinished} in Figure~\ref{code.isGameFinished} taken from the web game \emph{sudoku}\footnote{\url{http://www.dhtmlgoodies.com/scripts/game_sudoku/game_sudoku.html}}. This function checks the validity of a sudoku solution given by a user. First, we find the DOM element representing the game field (line~\ref{isGameFinished.getSudoku}). Then, we collect all its child DIVs (line~\ref{isGameFinished.getDivs}). Each DIV element corresponds to a sudoku input square (line~\ref{isGameFinished.if1.begin}) and consists of two SPANs. The first span contains a user input, whereas the second one, which is hidden by default, stores the expected value for that sudoku square. Inside of a for-loop in lines~\ref{isGameFinished.inFor.begin}-\ref{isGameFinished.outFor}, we iterate over the whole collection of input squares. Once, we find a square with two unequal SPAN values (line~\ref{isGameFinished.if2.begin}), the loop is terminated and the game is considered unfinished (line~\ref{isGameFinished.unfinished}) and finished otherwise.

Suppose now that we would like to unit test this function in isolation to the rest of the application. We aim to maximize the branch coverage as a common testing criteria~\cite{zhu1997software}. That is, for each individual branch of the function under test (FUT) \texttt{isGameFinished}, we need to construct a test input that covers the branch and leads to the normal termination of the FUT. The JS function takes \emph{explicit} input arguments (in our case none), but it can also accept \emph{implicit} arguments such as the DOM state. Henceforth, we assume that DOM is the only global variable that is allowed to be referred to in the body of the FUT. In general, in order to produce a complete test for our function we have to construct both an appropriate DOM and input arguments.

@@

Figure~\ref{fig.isGameFinished.tests} presents four tests that together provide the full coverage of the FUT. Let us look now at how following the structure of the FUT in Figure~\ref{code.isGameFinished}, we can construct the $T4$ test covering the $(10,11)$ branch. Line~\ref{isGameFinished.getSudoku} expects the input DOM to have an element with the id \textquotesingle\texttt{sudoku}\textquotesingle. In order to enter the for-loop in line~\ref{isGameFinished.inFor.begin}, that element should have at least one child \emph{DIV} tag. The first if-condition in line~\ref{isGameFinished.if1.begin} requires this $div$ to be of the class \textquotesingle\texttt{sudokuSquare}\textquotesingle\ and also have no background color. The second if-condition (line~\ref{isGameFinished.if2.begin}) expects that the $div$ consists of two \emph{SPAN} elements whose \emph{innerHtml} values are not equal. The resulting test $T4$ is shown in Figure~\ref{fig.isGameFinished.tests} and meets all the above conditions.

As we have just seen, even for such a relatively simple example, the construction procedure of a test input is far from trivial because it requires the deep understanding of the program semantics. In practice, web developers often have to deal with a large number of the JS frameworks and libraries freely available on the GitHub whose code is commonly untested and poorly documented. At the same time, unit testing is the most fundamental level of testing and the easiest to implement. It saves the developers from unexpected regressions and often documents the desired behavior of a program.

@@

\begin{figure}[t!]
  \centering 
  \begin{sequencediagram}[font=\scriptsize]

@@

Given a JS function $f$ and a set of branches $\mathbb{B}_f$, the ultimate goal of the testing framework is to find input arguments for $f$ that cover all branches from $\mathbb{B}_f$. Our JS testing framework, called \emph{JEDI}, is a client-server application which under the hood uses a genetic engine for test generation. The general workflow of the framework is depicted as a sequence diagram in Figure~\ref{fig.framework.architect}. The diagram consists of the four interacting components \User, \Server, \Validator, and \Client. The \Server is the key component of the  framework. It is responsible for the \emph{Initialization Phase} and the \emph{Genetic Phase}. Both phases will be covered in detail in Section~\ref{sub.sec.init.phase} and \ref{sub.sec.genetic.phase} respectively.

Each newly constructed HTML document has to be processed by the \Validator\footnote{\url{https://validator.w3.org/nu/}} which checks its syntactic correctness and reports back the set of identified errors. \Server rejects the document if any errors were revealed and generates a new one until the generated document has no more errors. \Client is a \emph{Node.js}\footnote{\url{https://nodejs.org/en/}} application that is mainly responsible for the execution of the JS code. We use the \emph{jsdom}\footnote{\url{https://github.com/tmpvar/jsdom}} library to model the virtual DOM and simulate the native browser API calls.

@@

Let us take a closer look at the sequence of interactions in Figure~\ref{fig.framework.architect}. As an input for our testing framework, \User has to provide the JS file \emph{fun.js} that contains the function under test (FUT) \emph{fun} annotated with a type signature. Additionally, \User specifies a configuration file \emph{config} that defines various parameters for random and genetic phases, logging etc. Experimentally, we have found an optimal set of input configurations that was used across all experimental objects in Section~\ref{sec.evaluation}. This data is processed by the \Server during the \emph{Initialization Phase}. As an output, we get back the following information about \emph{fun}: the control flow graph (\emph{cfg}), the complete list of its branches (\emph{branches}), the constant info (\emph{cinfo}), the instrumented version \emph{fun} (\emph{ifun}), and the signature (\emph{sig}). The \Server sends a \texttt{POST} request with \emph{ifun} and \emph{sig} to the \Client. Depending on the configured execution mode, the \Server either asks the \User to select a \emph{target} branch or covers all branches one by one in a fixed order. For each branch, our framework enters the \emph{Genetic Loop (Phase)}. Inside of this phase, the framework starts with a randomly generated population (\emph{Random Loop}), and continues to evolve that population (\emph{Fitness Loop}) until either a perfect entity is found or a certain termination criteria is achieved, e.g. the limit on the number of generation attempts. Each candidate in the population has to be evaluated in \emph{Evaluation Loop} by \Client. It matches the candidate $p$ with the type signature \emph{sig} and invokes the instrumented JS function on it: \emph{ifun(p)}. The execution trace along with some supplementary information about the execution are sent back to the \Server, where the \emph{fitness value} is computed. The new population is constructed by combining the results of the \emph{Crossover} and \emph{Mutation Loops}. If a new HTML document is produced somewhere in the genetic phase, it has to be first approved by the \Validator. At the end of the \emph{Fitness Loop}, the candidate with the ``best'' fitness value is reported back to the \User. Once all the branches are covered, the algorithm reaches the end. Below, we elaborate on the details of each individual phase of our test generation framework.

@@

During the \emph{Initialization Phase} we analyze the FUT in order to retrieve the data required for the subsequent genetic testing. The main steps of this phase are described in Algorithm~\ref{alg.init}. Given a FUT \emph{fun.js} as an input, we parse the function and its type signature in line~\ref{alg.init.parse.fun.sig}. In order to maximize the chances of generating the fittest candidate in the genetic phase, in line~\ref{alg.init.collect.info} we perform static analysis of the FUT with the purpose to collect information about constants. Such constants include the numeric and string literals, as well as the arguments to the following DOM API methods: \texttt{getElementsByTagName}, \texttt{getElementById}, \texttt{getElementsByClassName}, and \texttt{getElementsByName}
These methods provide the information about \emph{tags}, \emph{ids}, \emph{classes} and \emph{names} referred from the FUT. The next step is to normalize the FUT before applying any instrumentation (line~\ref{alg.init.transform.ast}), e.g. transform \emph{if-then} branches into \emph{if-then-else}. We instrument the FUT (line~\ref{alg.init.instr}) in order to collect the run-time data such as the execution trace, \emph{the approach level}, \emph{the branch and loop distances} (explained later), and the constants. Upon the \emph{ifun} execution this information is gathered and fed back to the fitness evaluation procedure. Finally, we construct the control-flow graph (line~\ref{alg.init.build.cfg}) and identify all branches of the FUT (line~\ref{alg.init.get.branches}).

@@

\begin{figure}[t!]
\setlength{\grammarparsep}{3pt}
\footnotesize
\begin{grammar}

<JS Type Annotation> ::= /*t <Signature> */

<Signature> ::= <DOM> : <Type Signature> | <Type Signature>

<DOM> ::= \texttt{dom}

@@

<Array Type> ::= [ <Primitive Type> ] | [ <Array Type> ]
\end{grammar}
\caption{Grammar of supported JS type annotations}
\label{fig.js.type.annot}
\end{figure}
One assumption made in the paper is the availability of type signatures for JS functions. We believe in the legitimacy of such a request due to the following observations. First, typed variants of JS such as \emph{TypeScript}\footnote{\url{https://www.typescriptlang.org/}} are rising in popularity. Second, there has been an active work on the JS type inference done both in academia~\cite{kristensen2017type} and industry, e.g. Facebook's tool \emph{Flow}\footnote{\url{https://flow.org/}}.

The grammar of the JS type annotations supported by our test generation framework is shown in Figure~\ref{fig.js.type.annot}. Each type annotation is a signature embedded into a specialized JS type comment (\texttt{/*t...*/}). We distinguish between two kinds of signatures: those which refer to the global DOM and ``ordinary'' JS type signatures. The functions with the former kind expect some \texttt{dom} argument to be passed as the first implicit parameter of the function call. Whereas ordinary JS signature is a non-empty sequence of the JS types. The type can be \emph{Void}, \emph{Primitive} (\texttt{bool}, \texttt{int}, \texttt{float}, \texttt{string}) or \emph{Array} (recursive homogeneous array type).

@@

Figure~\ref{code.isGameFinished.instr} presents the \texttt{isGameFinished} function after instrumentation. The global variable \emph{trace} in line~\ref{isGameFinished.instr.trace} is an array of labels corresponding to the sequence of executed program statements. All statements have to be traced because each one, in theory, can raise an exception. The \texttt{loopMap} map in line~\ref{isGameFinished.instr.loopMap} approximates the potential number of iterations of the respective for-loop. If no exact bound can be computed, we set it to one. The \texttt{branchDistance} map records the conditional distance~\cite{tracey1998automated} on the branch where the deviation from the target branch has happened.

@@

\begin{figure}[t]
  \begin{lstlisting}[style=htmlcssjs,language=JavaScript]
function isGameFinished() {
 trace.push(2); |c \label{isGameFinished.instr.trace} |c
 var obj = document.getElementById("sudoku");
 trace.push(3);
 var subDivs = obj.getElementsByTagName("DIV");
 trace.push(4);
 var allOk = true;
 loopMap[5] = function () { |c \label{isGameFinished.instr.loopMap} |c
   var no = 0;
   if (subDivs.length || subDivs.length == 0) {
     return Math.abs(no - subDivs.length);
   }
   return 1;
 }();
 trace.push(5);
 for (var no = 0; no < subDivs.length; no++) {
  trace.push(6);
  trace.push(7);
  if (subDivs[no].className.indexOf("sudokuSquare") >= 0 &&
      !subDivs[no].style.backgroundColor) {
   trace.push(8);
   branchDistance.push({ 
    label: 7, 
    distance: Math.min(
     abs(subDivs[no].className.indexOf("sudokuSquare"), 0) + _K_, 
     absZero(subDivs[no].style.backgroundColor))
   });
   trace.push(9);
   var spans = subDivs[no].getElementsByTagName("SPAN");
   trace.push(10);
   if (spans[0].innerHTML != spans[1].innerHTML) {
    trace.push(11);
    branchDistance.push({
     label: 10,
     distance: abs(spans[0].innerHTML, spans[1].innerHTML)
    });
    trace.push(12);
    allOk = false;
    break;
   } else {
     trace.push(14);
     branchDistance.push({label: 10,distance: _K_});
   }
  } else {
   trace.push(15);
   branchDistance.push({
    label: 7,
    distance: 
     abs(subDivs[no].className.indexOf("sudokuSquare"), 0) + 
     absNegZero(subDivs[no].style.backgroundColor)
   });
  }
 }
 trace.push(5);
 trace.push(-1);
}
\end{lstlisting}
  \caption{Instrumented version of \texttt{isGameFinished.js} }
  \label{code.isGameFinished.instr}
\end{figure}

@@

The heart of our test data generation framework is in the \emph{Genetic Phase}. Although, the main workflow of the underlying genetic algorithm (GA) is fairly standard~\cite{poli2008field}, it has been extended with the concept of  \emph{archive convergence} (Algorithm~\ref{alg.gen}). As an input, we take a configuration file \emph{config}, a target branch $b_n$ of the FUT (which we would like to cover), and a tuple of supplementary parameters computed by Algorithm~\ref{alg.init}. The output consists of the ``best'' fitness score $f^*$ achieved by the population candidate $p^*$.

First, we read the \emph{config} file (line~\ref{alg.gen.read.config}) and retrieve the following GA configuration parameters: a population size  ($size_P$), an archive size ($size_A$), the number of generations ($size_G$), crossover ($rate_C$) and mutation ($rate_M$) rates respectively, and a history window size ($size_H$). We start with the initial target $b_n$ (line~\ref{alg.gen.init.target}) and the iteration counter $i$ equal to 0 (line~\ref{alg.gen.iter.count}). We run the genetic search for the target branch (lines~\ref{alg.gen.converged.loop.begin}-\ref{alg.gen.converged.loop.end}) until either the history has not converged, or the generations limit has not been reached. The \emph{archive} variable stores the most fit population candidates observed so far, and the \emph{history} variable records all archives during the generation process. These variables both are empty at the beginning of the evolution (line~\ref{alg.gen.init.archive} and~\ref{alg.gen.init.history} respectively).

The initial population has $size_P$ elements, and it is randomly generated out of the constants from \emph{cinfo} (line~\ref{alg.gen.random.gen}). In the loop~\ref{alg.gen.loop.begin}-\ref{alg.gen.loop.end}, this population is evolved until either a perfect candidate has been found, or the history has converged, or the generation limit $size_G$ has been reached. But first, we need to score the current population (lines~\ref{alg.gen.eval.begin}-\ref{alg.gen.eval.end}). Each candidate $p$ of the population $pop$ represents a concrete substitution for the arguments of the FUT. Thus, it should to be coerced respectively to the type signature $sig$ (line~\ref{alg.gen.make.args}). Afterwards, the instrumented function \emph{ifun} can be called with the coerced arguments $args$ (line~\ref{alg.gen.call.ifun}). This gives us a 4-tuple consisting of an execution trace ($tr$), branch ($d_B$), loop distances ($d_L$), and the set of dynamic constants (\emph{dyncinfo}) collected at the run-time. The next step is actually computing the fitness value for $p$ (line~\ref{alg.gen.eval}). Collected dynamic constants \emph{dyncinfo} are merged with the existing constant pool (line~\ref{alg.gen.eval.end}). Eventually, the scored population is combined with the archive (line~\ref{alg.gen.get.combo}) in $combo$. And it is used for crossover (line~\ref{alg.gen.cross}) and mutation (line~\ref{alg.gen.mut}) steps in order to form the new population (line~\ref{alg.gen.new.pop}). Then, the scored population together with the current archive are sorted by fitness value, and the best $size_A$ elements form the new archive (line~\ref{alg.gen.new.archive}). The element at the top of the archive is the fittest candidate (line~\ref{alg.gen.archive.head}) of the current population. If $f^*$ is a \emph{perfect} candidate, the solution is found (line~\ref{alg.gen.loop.end}). Eventually, the tuple $(f^*, p^*)$ is returned by the algorithm (line~\ref{alg.gen.return}) as solution.

>>>>

\newcommand{\Server}{\emph{Server}\xspace}
\newcommand{\Client}{\emph{Client}\xspace}
\newcommand{\User}{\emph{User}\xspace}
\newcommand{\Validator}{\emph{Validator}\xspace}
\newcommand{\thenBr}{\text{then}}
\newcommand{\elseBr}{\text{else}}
\newcommand{\inFor}{\text{inFor}}
\newcommand{\inWhile}{\text{while}}
\newcommand{\un}[1]{\underline{#1}}
\newcommand{\Random}{\mathbb{R}}
\newcommand{\Genetic}{\mathbb{G}}
\newcommand{\RGenetic}{\mathbb{G}_{10}}
\newcommand{\Confix}{\emph{Confix}}
\newcommand{\Jalangi}{\emph{Jalangi}}
\newcommand{\Jedi}{\emph{JEDI}}

@@

\begin{abstract}
The popularity of JavaScript is continuously growing. Together with HTML and CSS, it forms the core technology stack for web development. Because of the dynamic nature and complex interplay with HTML, JS applications are often error-prone and vulnerable. Despite the active research efforts to devise intricate static and dynamic analyses to facilitate JS testing, the problem of test data generation for JS applications interacting with the DOM has not yet been addressed. In this paper, we present a novel \textbf{J}avascript \textbf{E}volution-based testing framework with \textbf{D}OM as an \textbf{I}nput, called \emph{JEDI}. In order to reach a target branch, it applies genetic search for relevant input parameters of the JS function including the global DOM state. We conducted an empirical evaluation to study the effectiveness and efficiency of our testing framework. It shows that the \emph{genetic with restart} algorithm, proposed in this paper, is able to achieve complete branch coverage for all experimental subjects, taking on average 19 seconds per branch.
\end{abstract}

@@

\begin{figure}[t]
  \begin{lstlisting}[style=htmlcssjs,language=JavaScript]
/*t dom */
function isGameFinished() {
  var obj = document.getElementById('sudoku'); |c \label{isGameFinished.getSudoku} |c
  var subDivs = obj.getElementsByTagName('DIV'); |c \label{isGameFinished.getDivs} |c
  var allOk = true;
  for (var no = 0; no < subDivs.length; no++) { |c \label{isGameFinished.inFor.begin} |c
    if (subDivs[no].className.indexOf('square') >= 0 |c \label{isGameFinished.if1.begin} |c 
        && !subDivs[no].style.backgroundColor) { 
      var spans = subDivs[no].getElementsByTagName('SPAN');
      if (spans[0].innerHTML != spans[1].innerHTML) { |c \label{isGameFinished.if2.begin} |c
        allOk = false; //target |c \label{isGameFinished.unfinished} |c
        break;}}}} |c \label{isGameFinished.outFor} |c
\end{lstlisting}
  \caption{JS function \texttt{isGameFinished.js} from sudoku}
  \label{code.isGameFinished}
\end{figure}

@@

<!-- T3: (7,8), (10,14) -->   <!-- T4: (7,8), (10,11) -->
<html>                        <html>
 <body>                        <body>
  <div id='sudoku'>             <div id='sudoku'>
   <div class='square'>          <div class='square'>
    <span></span>                 <span></span> 
    <span></span>                 <span>TEST</span>
   </div>                        </div> 
  </div>                        </div>
 </body>                       </body>
</html>                       </html>
  \end{lstlisting}
  \caption{Input DOM arguments for \texttt{isGameFinished.js}}
  \label{fig.isGameFinished.tests}
\end{figure}

Let us consider the JS function \texttt{isGameFinished} in Figure~\ref{code.isGameFinished} taken from the web game \emph{sudoku}~\cite{sudoku}. This function validates a sudoku solution given by a user. First, it finds a DOM element representing the game field (line~\ref{isGameFinished.getSudoku}). Then, it collects all the child \emph{divs} (line~\ref{isGameFinished.getDivs}). Each \emph{div} element corresponds to a sudoku input square (line~\ref{isGameFinished.if1.begin}), and consists of two \emph{spans}. The first \emph{span} contains a user input, whereas the second one (hidden by default) stores an expected value for that sudoku square. Inside of a for-loop in lines~\ref{isGameFinished.inFor.begin}-\ref{isGameFinished.outFor}, we iterate over the whole collection of input squares. Once, we find a square with two unequal values (line~\ref{isGameFinished.if2.begin}), the loop is terminated and the game is considered unfinished (line~\ref{isGameFinished.unfinished}) and finished otherwise.

Suppose now that we would like to unit test this function in isolation to the rest of the application. We aim to maximize the branch coverage as a common testing criteria~\cite{zhu1997software}. That is, for each individual branch of the function under test (FUT) \texttt{isGameFinished}, we need to construct a test input that covers the branch and leads to the normal termination of the FUT. A JS function can take \emph{explicit} input arguments (in our case none), but it can also accept \emph{implicit} arguments such as a DOM state. In general, in order to produce a complete test for our function we have to construct both an appropriate DOM and input arguments.

@@

Figure~\ref{fig.isGameFinished.tests} presents four tests that together provide the full coverage of the FUT. Let us take a look at how to construct the \texttt{T4} test, which covers the branch $(10,11)$, by following the structure of the FUT. Line~\ref{isGameFinished.getSudoku} expects the input DOM to have an element with an id \textquotesingle\texttt{sudoku}\textquotesingle. In order to enter the for-loop in line~\ref{isGameFinished.inFor.begin}, that element should have at least one child \emph{div} tag. The first if-condition in line~\ref{isGameFinished.if1.begin} requires this \emph{div} to be of the class \textquotesingle\texttt{square}\textquotesingle\ and also have no background color. The second if-condition (line~\ref{isGameFinished.if2.begin}) expects that the \emph{div} consists of two \emph{span} elements whose \emph{innerHtml} values are not equal. The resulting test \texttt{T4} in Figure~\ref{fig.isGameFinished.tests} meets all the above conditions.

As we have just observed, even for such a relatively simple example, the test data generation can be far from trivial procedure because it requires the deep understanding of the program semantics. In practice, web developers often have to deal with a large number of the JS frameworks and libraries freely available on the GitHub whose code is commonly untested and poorly documented. At the same time, unit testing is the most fundamental level of testing and the easiest to implement. It saves the developers from unexpected regressions and often documents the desired behavior of a program.

@@

\begin{figure}[t!]
  \centering 
\tikzset{
  % add this style to all tikzpicture environments
  every picture/.append style={
    % enable scaling of nodes
    transform shape,
    % set scale factor
    scale=1
  }
}
  \begin{sequencediagram}[font=\scriptsize]

@@

Given a JS function $f$ and a set of branches $\mathbb{B}_f$, the ultimate goal of the testing framework is to find input arguments for $f$ that cover all branches from $\mathbb{B}_f$. Our JS testing framework, called \emph{JEDI}, is a client-server application, which uses under the hood a genetic engine, for test generation. The general workflow of the framework is depicted as a sequence diagram in Figure~\ref{fig.framework.architect}. The diagram consists of the four interacting components \User, \Server, \Validator, and \Client. The \Server is the key component of the  framework, responsible for the \emph{Initialization} and \emph{Genetic Phases}, covered in detail in Section~\ref{sub.sec.init.phase} and \ref{sub.sec.genetic.phase} respectively.

Each newly constructed HTML document has to be assessed by the \Validator~\cite{htmlvalidator} that checks its syntactic correctness and reports back any identified errors. The document generation process is repeated by the \Server until there are no errors revealed. \Client is a \emph{Node.js}~\cite{nodejs} application that is mainly responsible for the execution of the JS code. We use the \emph{jsdom}~\cite{jsdom} library to model the virtual DOM and simulate the native browser API calls.

@@

Let us take a closer look at the sequence of interactions in Figure~\ref{fig.framework.architect}. As an input for our testing framework, \User has to provide a JS file \emph{fun.js} that contains a FUT \emph{fun} annotated with the type signature. Additionally, \User specifies a configuration file \emph{config} that sets various control parameters for the random and genetic phases, logging etc. Experimentally, we have found an optimal configuration that was consistently used in the experimental evaluation in Section~\ref{sec.evaluation}. Once all the data are processed by the \Server in \emph{Initialization Phase}, we get back the following information about \emph{fun}: the control flow graph (\emph{cfg}), the complete list of its branches (\emph{branches}), the constant info (\emph{cinfo}), the instrumented version \emph{fun} (\emph{ifun}), and the signature (\emph{sig}). Then, the \Server sends a \texttt{POST} request with \emph{ifun} and \emph{sig} to the \Client. Depending on the configured execution mode, the \Server either asks the \User to select a \emph{target} branch or covers all branches one by one in a fixed order. For each branch, our framework triggers the \emph{Genetic Loop (Phase)}. Inside of this phase, the framework begins with a randomly generated population (\emph{Random Loop}), and continues to evolve that population (\emph{Fitness Loop}) until either a perfect entity is found or a termination criteria is achieved, e.g. the limit on the number of generation attempts. Any HTML document produced during the genetic phase has to be first approved by the \Validator. Each candidate in the population has to be evaluated in \emph{Evaluation Loop} by \Client. It matches the candidate solution $p$ with the type signature \emph{sig}, and invokes the instrumented JS function as \emph{ifun(p)}. The execution trace along with some auxiliary information is sent back to the \Server, where the final \emph{fitness value} is computed. A new population is constructed by combining the results of the \emph{Crossover} and \emph{Mutation Loops}. At the end of the \emph{Fitness Loop}, the candidate with the ``best'' fitness value is reported back to the \User. Once all the branches are covered, the algorithm terminates. Below, we elaborate on the details of each individual phase of our test generation framework.

@@

During the \emph{Initialization Phase} we analyze the FUT in order to facilitate the subsequent genetic testing. The main steps of this phase are described in Algorithm~\ref{alg.init}. Given a FUT \emph{fun.js} as an input, we parse the function and its type signature in line~\ref{alg.init.parse.fun.sig}. In order to maximize the chances of generating the fittest candidate, in line~\ref{alg.init.collect.info} we perform static analysis of the FUT with the purpose to collect information about constants, e.g. numeric and string literals. Additionally, we monitor the calls to such DOM API methods as \texttt{getElementsByTagName}, \texttt{getElementById}, and \texttt{getElementsByClassName}. These methods provide the information about \emph{tags}, \emph{ids}, \emph{classes} and \emph{names} referred from the FUT. The next step is to normalize the FUT before applying any instrumentation (line~\ref{alg.init.transform.ast}), e.g. transform \emph{if-then} branches into \emph{if-then-else}. We instrument the FUT (line~\ref{alg.init.instr}) for run-time data collection about the execution trace, \emph{the approach level}, \emph{the branch and loop distances} (explained later), and the constants. Upon the \emph{ifun} execution this information is gathered and fed back to the fitness evaluation procedure. Finally, we construct the control-flow graph (line~\ref{alg.init.build.cfg}) and identify all branches of the FUT (line~\ref{alg.init.get.branches}).

@@

\begin{figure}[t!]
\setlength{\grammarparsep}{3pt}
\footnotesize
\begin{grammar}
<JS Type Annotation> ::= /*t <Signature> */

<Signature> ::= \texttt{dom} : <Type Signature> | <Type Signature>

@@

<Array Type> ::= [ <Primitive Type> ] | [ <Array Type> ]
\end{grammar}
\caption{Grammar of supported JS type annotations}
\label{fig.js.type.annot}
\end{figure}

We assume the FUT is equipped with the type signature. The legitimacy of such a requirement can be confirmed by the popularity of \emph{TypeScript}~\cite{typescript}, and advances in the JS type inference \emph{Flow}~\cite{flow}. The supported grammar of type annotations is shown in Figure~\ref{fig.js.type.annot}. Each type annotation is a signature embedded into a specialized JS type comment (\texttt{/*t...*/}). We distinguish between two kinds of signatures: with the reference to the global DOM and ``ordinary'' JS type signatures. The former kind expects the \texttt{dom} terminal to be in place of the first parameter of the function call. Whereas an ordinary JS signature is a non-empty sequence of the JS types. The type can be \emph{Void}, \emph{Primitive} (\texttt{bool}, \texttt{int}, \texttt{float}, \texttt{string}) or \emph{Array} (recursive homogeneous array type).

@@

Figure~\ref{code.isGameFinished.instr} presents the \texttt{isGameFinished} function after instrumentation. The global variable \texttt{trace} in line~\ref{isGameFinished.instr.trace} records the sequence of executed statements of the FUT. We have to trace all the statements because each one, in theory, can rise an exception. The \texttt{loopMap} variable in line~\ref{isGameFinished.instr.loopMap} captures the upper bound for the number for-loop iterations. If the exact bound is unknown, it is set to one. The \texttt{branchDistance} variable, computed for the \emph{problem node}, contains the distance~\cite{tracey1998automated} from the target branch.

@@

\begin{figure}[t]
  \begin{lstlisting}[style=htmlcssjs,language=JavaScript]
function isGameFinished() {
 trace.push(1);var obj = document.getElementById("sudoku"); |c \label{isGameFinished.instr.trace} |c
 trace.push(2);var subDivs = obj.getElementsByTagName("DIV");
 var allOk = true;
 loopMap[3] = function () { |c \label{isGameFinished.instr.loopMap} |c
   var no = 0;
   if (subDivs.length || subDivs.length == 0) {
     return Math.abs(no - subDivs.length);
   }; return 1}();
 trace.push(3);
 for (var no = 0; no < subDivs.length; no++) {
  trace.push(4);
  if (subDivs[no].className.indexOf("square") >= 0 &&
      !subDivs[no].style.backgroundColor) {
   trace.push(5); branchDistance.push({ label: 3, 
    distance: Math.min(
     abs(subDivs[no].className.indexOf("square"),0)+_K_ 
     absZero(subDivs[no].style.backgroundColor))
   });
   trace.push(6);var spans=subDivs[no].getElementsByTagName("SPAN");
   trace.push(7);
   if (spans[0].innerHTML != spans[1].innerHTML) {
    trace.push(8); branchDistance.push({ label: 7, 
     distance: abs(spans[0].innerHTML, spans[1].innerHTML)
    });
    allOk = false; break;
   } else {
   trace.push(9);branchDistance.push({label:7,distance:_K_});
   }
  } else {
   trace.push(10); branchDistance.push({ label: 3,
    distance: 
     abs(subDivs[no].className.indexOf("square"), 0) + 
     absNegZero(subDivs[no].style.backgroundColor)
   });}}
 trace.push(-1);
}
\end{lstlisting}
  \caption{Instrumented version of \texttt{isGameFinished.js} }
  \label{code.isGameFinished.instr}
\end{figure}

@@

The heart of our test data generation framework is in the \emph{Genetic Phase}. Although, the main workflow of the underlying genetic algorithm (GA) is fairly standard~\cite{poli2008field}, it has been extended with the concept of \emph{archive convergence} (Algorithm~\ref{alg.gen}). As an input, we specify a configuration file \emph{config}, a target branch $b_n$ of the FUT, and a tuple of supplementary parameters computed by Algorithm~\ref{alg.init}. The output consists of the ``best'' fitness score $f^*$ achieved by the population candidate $p^*$.

First, we read the \emph{config} file (line~\ref{alg.gen.read.config}) to retrieve the following GA settings: a population size  ($size_P$), an archive size ($size_A$), the number of generations ($size_G$), crossover ($rate_C$) and mutation ($rate_M$) rates respectively, and a history window size ($size_H$). We start with the initial target $b_n$ (line~\ref{alg.gen.init.target}) and the iteration counter $i$ set to 0 (line~\ref{alg.gen.iter.count}). We run the genetic search for the target branch (lines~\ref{alg.gen.converged.loop.begin}-\ref{alg.gen.converged.loop.end}) until either the history has not converged, or the generations limit has not been exceeded. The \emph{archive} variable stores the most fit population candidates observed so far, and the \emph{history} variable records all the archives during the generation process. The two variables are empty at the beginning of the evolution (line~\ref{alg.gen.init.archive} and~\ref{alg.gen.init.history} respectively).

The initial population of $size_P$ is randomly generated out of the constants from \emph{cinfo} (line~\ref{alg.gen.random.gen}). In the loop~\ref{alg.gen.loop.begin}-\ref{alg.gen.loop.end}, this population is evolved until either a perfect candidate is found, or the history is converged, or the generation limit $size_G$ is reached. But first, we need to score the current population (lines~\ref{alg.gen.eval.begin}-\ref{alg.gen.eval.end}). Each candidate $p$ of the population $pop$ represents a concrete substitution for the arguments of the FUT. Thus, it should to be coerced respectively to the type signature $sig$ (line~\ref{alg.gen.make.args}). Afterwards, the instrumented function \emph{ifun} can be called with the coerced arguments $args$ (line~\ref{alg.gen.call.ifun}). This gives us a 4-tuple consisting of an execution trace ($tr$), branch ($d_B$), loop distances ($d_L$), and the set of dynamic constants (\emph{dyncinfo}) collected at the run-time. The next step is actually computing the fitness value for $p$ (line~\ref{alg.gen.eval}). Collected dynamic constants \emph{dyncinfo} are merged into the existing constant pool (line~\ref{alg.gen.eval.end}). Eventually, the scored population is combined with the archive (line~\ref{alg.gen.get.combo}) in $combo$. It is used for crossover (line~\ref{alg.gen.cross}) and mutation (line~\ref{alg.gen.mut}) steps in order to form the new population (line~\ref{alg.gen.new.pop}). Then, the scored population together with the current archive are sorted by fitness value, and the best $size_A$ elements form the new archive (line~\ref{alg.gen.new.archive}). The element at the top of the archive is the fittest candidate (line~\ref{alg.gen.archive.head}) of the current population. If $f^*$ is a \emph{perfect} candidate, the solution is found (line~\ref{alg.gen.loop.end}). Eventually, the tuple $(f^*, p^*)$ is returned by the algorithm (line~\ref{alg.gen.return}) as solution.
